Do researchers preferentially collaborate with same-gendered colleagues?

### 2018-05-22 manuscript/Manuscript_Claire.Rmd
Manuscript_Claire.Rmd

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
author:
- Luke Holman* and Claire Morandin
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Conference gender ratios} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and that many female researchers feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping redress gender imbalance in the academic workforce, increasing women’s access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this ‘gender homophily’ is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

# Introduction
Women are substantially underrepresented in many branches of the workforce in science, technology, engineering, mathematics, and medicine (STEMM), and they face additional challenges and inequities relative to men [e.g. @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions (Wutte 2007, Reuben, Sapienza & Zingales, 2014) with lower salaries (Trower and Chait 2002, Umback 2007), receive less grant money (Hosek et al. 2005, OER 2005), are promoted more slowly (Hopkins et al. 2013; Long et al. 1993, Rosenfeld 1991, Zuckerman 1987), and are allocated fewer resources (European Commission 2009) and less research funding (Feldt 1986, Stack 2004, Larivière et al. 2011) than men. Experimental studies have demonstrated that researchers regard women's achievements less favourably than identical achievements by men (Moss et al. 2012; Knobloch et al. 2013).

Because publishing, networking and collaboration are all instrumental to academic career advancement (Lee and Boheman 2005, Hopkins et al. 2012), numerous studies have tested for gender differences in these areas (reviewed in Table S1 in Holman et al. 2018; see also XXX). For example, studies have shown that relative to men, women tend to be less involved in international collaboration (Lemoine 1995; Lewison 2001; Webster 2001; Larivière et al. 2011, Bozeman and Corley 2004), collaborate less with researchers from their own university department (Websters 2001), have less prestigious collaborations (Long 1990), and fewer collaborations in general (Fuchs Von Stebut and Allmendinger 2001). Possible explanations for the smaller average professional networks of women include a greater average amount of family obligations (Long 1990, Wright et al. 2003, Reskin 1978), lower participation of women in international conferences (Lewison 2001, Eisen 2014, Martin 2014), and reduced receipt of research travel funds by women (Bozeman and Corley 2004).

A high, steadily increasing proportion of research papers is written by more than one author (West et al. 2013), making collaboration a key predictor of publication output, and in turn, a key predictor of career prospects (Cohn et al 2014, Jordan et al. 2008, Tower et al. 2007). Additionally, empirical studies suggest that mixed-gender (Campbell et al. 2013) or otherwise diverse (Hong and Page 2004) teams can produce better results on collaborative tasks. For reasons such as these, several studies have tested for gender differences in collaboration frequency or pattern by examining the author lists of published research. To our knowledge, every study of this question has concluded that men publish with other men, and women with women, more often than expected if collaborators assort randomly with regards to gender (Ferber and Teiman, 1980; McDowell and Smith 1992; Ghiasi et al. 2015; Crow and Smykla 2015, Zettler et al. 2016, Fahmy and Young, 2016, Jadidi et al. 2017; Teele and Thelen 2017, Araujo and Fontainha 2017a,b). This preference for same-gendered colleagues is often termed ‘gender homophily’.

However, we hypothesise that these studies of gender homophily were hindered by a sampling issue that we term the Wahlund effect (\autoref{wahlund_plot}), making the true extent of homophily difficult to ascertain from present data. Essentially, whenever samples are taken from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can make it appear as though authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if researchers preferentially select opposite-gendered colleagues. For example, a sample of bioinformatics and cell biology papers might contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because more women work in cell biology than in bioinformatics [@Holman_2018]. We name this issue after the Wahlund effect in population genetics (Wahlund 1928), whereby spatial differences in allele frequencies create an excess of homozygotes across the whole population, even if alleles assort randomly within each subpopulation.

@@

# Methods
## Obtaining author lists and assigning gender
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every single article indexed on PubMed, from the oldest articles up to the present (which was August 2016), and attempting to assign a gender to each author based on their given name, using the web service genderize.io. Each journal was assigned to one of 107 scientific disciplines (e.g. “Nephrology” or “Cell Biology”), using PubMed’s journal categorisations where available. Because the present study focuses on co-authorship, all single-author papers were removed from the dataset. To simplify the statistical analysis, we also discarded all papers for which we could not determine the gender of every author with at least 95% certainty. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published 0-1 or 10-11 years before the PubMed data were collected (i.e. 20 August 2016). Lastly, we excluded journals with fewer than 50 suitable papers. This left us with a dataset from August 2015 - August 2016 containing `r sample.size$n2015[3]` papers with `r sample.size$n2015[4]` authorships, and a dataset from August 2005 - August 2006 containing `r sample.size$n2005[3]` papers and `r sample.size$n2005[4]` authorships. These papers came from `r sample.size$n2015[2]` and `r sample.size$n2005[2]` journals respectively, which were grouped into `r sample.size$n2015[1]` and `r sample.size$n2005[1]` research disciplines. There was a median of `r sample.size$n2015[5]` (`r sample.size$n2005[5]`) papers per journal, `r sample.size$n2015[6]` (`r sample.size$n2005[6]`) authors per journal, and `r sample.size$n2015[7]` (`r sample.size$n2005[7]`) authors per paper (the first number is for 2015-6, and the bracketed number for 2005-6).

@@

To estimate $\alpha$ for a particular subset of the scientific literature, we estimated $p$ as the average proportion of men’s co-authors who are men (averaged across all papers with at least one man author), and $q$ as the average proportion of women’s co-authors who are men (averaged across all papers with at least one woman author). To estimate the 95% confidence intervals on $\alpha$ for a given set of $n$ papers, we sampled $n$ papers with replacement 1000 times, estimated $\alpha$ on each sample, and recorded the 95% quantiles of the 1000 resulting estimates.

@@

Our test assumes that the expected value of $\alpha$ is zero if authors randomly assort, but for small datasets this assumption is not always true (as pointed out by Carl T. Bergstrom in a blog post, http://www.eigenfactor.org/gender/assortativity/note_to_eisen.rtf). To borrow Prof. Bergstrom’s example, consider a small research specialty comprising just two men and two women researchers, who have together produced six two-author papers: one paper in each of the six possible two-author combinations. For these six papers, $\alpha$ = $-\frac{1}{3}$, even though same- and opposite-gendered coauthors were selected in equal proportion to their frequency in the pool of possible collaborators. To control for the fact that the null expectation for $\alpha$ is not necessarily zero (particularly for very small datasets), we devised an adjusted version of the coefficient of homophily, which we term $\alpha'$. Every time we calculated $\alpha$ for a set of papers, we also determined the expected value of $\alpha$ under the null hypothesis that authors assort randomly with respect to gender. This was accomplished by randomly shuffling authors across papers 1000 times, recalculating $\alpha$, and taking the median. We then calculated $\alpha'$ by subtracting the null expectation for $\alpha$ from the observed value. We also used the null-simulated $\alpha$ values to calculate a two-tailed p-value for the observed value of $\alpha$; the p-value was defined as the proportion of null simulations for which $|\alpha_{null}| > |\alpha_{obs}|$. We applied false discovery rate correction to each set of p-values to account for multiple testing (Benjamini and Hochberg 1995).

@@

## Minimising the Wahlund effect: discipline and time period
As discussed in the Introduction, the Wahlund effect can give the appearance of gender homophily (reflected in our study as $\alpha'$ > 0) whenever one lumps together different ‘subpopulations’ of academic papers that differ in their author gender ratios (Figure 1).

@@

## Statistical analysis
Previous authors (e.g. Bonham & Stefan 2017) have hypothesised that senior scientists preferentially recruit staff and students of the same gender, and/or that junior researchers preferentially select same-gendered mentors. In the majority of disciplines, authorship conventions mean that the first-listed author is often an early-career researcher, while the author listed last is more likely to be a senior researcher leading a research team (Wren et al. 2007). Assuming that senior researchers are the main drivers of homophily and that there are enough papers with three or more authors, we predict that the last author’s gender will be the strongest predictor of the remaining authors’ genders (i.e. the gender of the last author will be more salient than that of the first author, or any other authorship position). This is because the first author’s gender would simply be an imperfect correlate of the true causal effect, while the last author’s gender would be the causal effect itself.

@@

$\alpha'$ was significantly higher in the sample from 2015-16 relative to 2005-6, but the difference in means was small (Figure S3; Effect of the fixed factor ‘time period’ in a linear mixed model of the data for all author positions: Cohen’s $d$ = 0.09±0.04, $t_{948}$ = 2.51, p = 0.012).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (Figure S4; Effect of the fixed factor ‘Authorship position’ in a linear mixed model: Cohen’s $d$ = 0.07±0.02, $t_{1988}$ = 4.48, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors’ genders than the gender of the last author, i.e. the opposite of our prediction based on the hypothesis that senior scientists are causally responsible for homophily.

## Possible differences in the strength of homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was 0.12±0.02 for Urology journals, and 0.03±0.01 for Veterinary Medicine journals; \autoref{alpha_histograms}, Table S3). However, there was little evidence for strong differences in $\alpha'$ between disciplines: the random factor ‘discipline’ explained at most 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). This implies that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

There was no indication that journals publishing on a wide range of topics have higher $\alpha'$ values than more specialised journals, due to the Wahlund effect. For example, the journal category “Multidisciplinary” - which includes journals like _PLoS ONE_, _Nature_, _Science_, and _PNAS_ - did not have notably elevated $\alpha'$ (\autoref{alpha_histograms}). This result suggests that our estimates of homophily, and estimates from earlier studies, are not notably inflated by the presence of disparate research topics (with variable author gender ratios) being published within individual journals.

@@

# Discussion
We found evidence that researchers tend to publish with same-gendered coauthors, despite implementing stringent controls for ‘spurious’ homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our data therefore reaffirm earlier conclusions (Ferber and Teiman 1980; McDowell and Smith 1992; Crow and Smykla 2015; Ghiasi et al. 2015; Zettler et al. 2016; Fahmy and Young 2016; Jadihi et al. 2017; Araujo and Fontainha 2017a; Teele and Thelen 2017) that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Many journals had $\alpha'$ values over 0.1, meaning that the gender ratio of the average man's coauthors is skewed towards men by >10%, relative to the gender ratio of a woman's coauthors. Few journals had $\alpha'$ values below zero, and fewer still had significantly negative $\alpha'$ values, suggesting that it is comparatively rare for researchers to preferentially select opposite-gendered collaborators.

We found no evidence that gender homophily is restricted to particular disciplines: instead, there was evidence of homophily across many diverse research disciplines. We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. This result suggests that recent increases in the number of women working in STEMM, as well as recent efforts to recruit, retain, and celebrate women in STEMM, have co-occurred with rise in the frequency of same-gender collaborations, not a fall. We speculate that this trend might result, at least in part, from the increasing number of women working in senior positions in STEMM over the last 10 years (DATA?). Because many co-authorships involve a junior and a senior researcher (e.g. a student and professor), and the gender ratio is typically more skewed among senior staff (Shaw and Stanton ref), there will be more opposite-gender collaborations than expected if the gender ratio were the same at both career stages. Thus, the observed increase in homophily might be due to the greater numbers of women working in leading roles.

Gender homophily was slightly stronger in research disciplines with a female-biased gender ratio, as compared to male-biased or gender-balanced fields. One interpretation of this result is that men are more likely to seek out other men in fields in which men are in the minority, relative to the homophily shown by women in fields in which women are in the minority.

We also found that journals with a high impact factor (relative to other journals in the same research discipline) tended to have weaker gender homophily than did low-impact journals. If one accepts the contentious claim that papers published in high-impact journals tend to be of higher quality than those in low-impact journals [e.g. @Garfield_2006], then this result provides correlational support for the hypothesis that mixed-gender teams produce better research than single-gender teams (Campbell et al. 2013).

Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research.

Academic publishing is one of the most (if not the most) important tool for a researcher to display their work, and is frequently used as a success indicator for being promoted, hired as a faculty member, or being invited as plenary speaker. Our finding that researchers are more likely to publish with same-gendered coauthors implies that co-authorship listing is not a random process, but instead creates additional gender imbalance. Homophily could make it harder for women to find collaborators in disciplines where women are in the minority [i.e. almost all STEMM fields; @Holman_2018]. Because women make up for as little as 30% of the STEMM scientist community (REF), they are therefore less likely to be picked as coauthors, and, as a result, gender homophily creates further gender imbalance and disadvantages for women. Men tend to publish more in high impact journals (REF) as first author, and generate more citations than papers lead by first author women (REF). As a consequence, gender homophily is also likely to contribute to lower publication rates for women (McDowell and Smith 1992), and lower scientific success (REF).

Why does gender homophily exists in science? Collaboration often leads to co-authorship publications, citations and additional professional recognitions, therefore first author should always be considering collaborators that are most likely to enhance their productivity and in the long term also their academic success, rather than pick co-authors according to their gender. First of all, it is likely that decades ago, the probability to meet a woman colleague working on the same area of expertise was slimmer than now. However, the number of women in STEM is increasing, and it is not likely to be an obstacle to collaboration nowadays.

# Acknowledgements
We are grateful to XXXX.

>>>>

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
author:
- Luke Holman* and Claire Morandin
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Conference gender ratios} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and that many female researchers feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping redress gender imbalance in the academic workforce, increasing women???s access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this ???gender homophily??? is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

# Introduction
Women are substantially underrepresented in many branches of the workforce in science, technology, engineering, mathematics, and medicine (STEMM), and they face additional challenges and inequities relative to men [e.g. @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions [@Wutte_2007; @Reuben_2014] with lower salaries [@Trower_2002; @Umbach_2007], receive less grant money [@Hosek_2005; @OER_2008], are promoted more slowly [@Zuckerman_1987; @Rosenfeld_1991; @Long_1993; @Hopkins_2013], and are allocated fewer resources [@ODorchai_2009] and less research funding [@Feldt_1986, @Stack_2004, @Lariviere_2011] than men. Experimental studies have demonstrated that researchers regard women's achievements less favourably than identical achievements by men [@Moss_2012; @Knobloch_2013].

Because publishing, networking and collaboration are all instrumental to scientific productivity and thus, academic career advancement [@Lee_2005; @Wuchty_2007;@Abramo_2009;@Lariviere_2015], numerous studies have tested for gender differences in these areas [reviewed in Table S1 in @Holman_2018; see also @Bozeman_2011; @Abramo_2013; @Badar_2013; @Long_1992]. For example, studies have shown that relative to men, women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less with researchers from their own university department [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in general [@Fuchs_2001]. Possible explanations for the smaller average professional networks of women include a greater average amount of family obligations [@Reskin_1978; @Long_1990; @Wright_2003], lower participation of women in international conferences [@Lewison_2001; Eisen 2014; @Martin_2014], and reduced receipt of research travel funds by women [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and in turn, a key predictor of career prospects [@Tower_2007; @Jordan_2008; @Cohn_2014]. Additionally, empirical studies suggest that mixed-gender [@Bear_2011; @Campbell_2013] or otherwise diverse [@Hong_2004] teams can produce better results on collaborative tasks. For reasons such as these, several studies have tested for gender differences in collaboration frequency or pattern by examining the author lists of published research. To our knowledge, every study of this question has concluded that men publish with other men, and women with women, more often than expected if collaborators assort randomly with regards to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015;@Fahmy_2017; @Zettler_2017;@Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This preference for same-gendered colleagues is often termed ???gender homophily???.

However, we hypothesise that these studies of gender homophily were hindered by a sampling issue that we term the Wahlund effect (\autoref{wahlund_plot}), making the true extent of homophily difficult to ascertain from present data. Essentially, whenever samples are taken from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can make it appear as though authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if researchers preferentially select opposite-gendered colleagues. For example, a sample of bioinformatics and cell biology papers might contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because more women work in cell biology than in bioinformatics [@Holman_2018]. We name this issue after the Wahlund effect in population genetics [@Wahlund_1928], whereby spatial differences in allele frequencies create an excess of homozygotes across the whole population, even if alleles assort randomly within each subpopulation.

@@

# Methods
## Obtaining author lists and assigning gender
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every single article indexed on PubMed, from the oldest articles up to the present (which was August 2016), and attempting to assign a gender to each author based on their given name, using the web service genderize.io. Each journal was assigned to one of 107 scientific disciplines (e.g. ???Nephrology??? or ???Cell Biology???), using PubMed???s journal categorisations where available. Because the present study focuses on co-authorship, all single-author papers were removed from the dataset. To simplify the statistical analysis, we also discarded all papers for which we could not determine the gender of every author with at least 95% certainty. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published 0-1 or 10-11 years before the PubMed data were collected (i.e. 20 August 2016). Lastly, we excluded journals with fewer than 50 suitable papers. This left us with a dataset from August 2015 - August 2016 containing `r sample.size$n2015[3]` papers with `r sample.size$n2015[4]` authorships, and a dataset from August 2005 - August 2006 containing `r sample.size$n2005[3]` papers and `r sample.size$n2005[4]` authorships. These papers came from `r sample.size$n2015[2]` and `r sample.size$n2005[2]` journals respectively, which were grouped into `r sample.size$n2015[1]` and `r sample.size$n2005[1]` research disciplines. There was a median of `r sample.size$n2015[5]` (`r sample.size$n2005[5]`) papers per journal, `r sample.size$n2015[6]` (`r sample.size$n2005[6]`) authors per journal, and `r sample.size$n2015[7]` (`r sample.size$n2005[7]`) authors per paper (the first number is for 2015-6, and the bracketed number for 2005-6).

@@

To estimate $\alpha$ for a particular subset of the scientific literature, we estimated $p$ as the average proportion of men???s co-authors who are men (averaged across all papers with at least one man author), and $q$ as the average proportion of women???s co-authors who are men (averaged across all papers with at least one woman author). To estimate the 95% confidence intervals on $\alpha$ for a given set of $n$ papers, we sampled $n$ papers with replacement 1000 times, estimated $\alpha$ on each sample, and recorded the 95% quantiles of the 1000 resulting estimates.

@@

Our test assumes that the expected value of $\alpha$ is zero if authors randomly assort, but for small datasets this assumption is not always true (as pointed out by Carl T. Bergstrom in a blog post, http://www.eigenfactor.org/gender/assortativity/note_to_eisen.rtf). To borrow Prof. Bergstrom???s example, consider a small research specialty comprising just two men and two women researchers, who have together produced six two-author papers: one paper in each of the six possible two-author combinations. For these six papers, $\alpha$ = $-\frac{1}{3}$, even though same- and opposite-gendered coauthors were selected in equal proportion to their frequency in the pool of possible collaborators. To control for the fact that the null expectation for $\alpha$ is not necessarily zero (particularly for very small datasets), we devised an adjusted version of the coefficient of homophily, which we term $\alpha'$. Every time we calculated $\alpha$ for a set of papers, we also determined the expected value of $\alpha$ under the null hypothesis that authors assort randomly with respect to gender. This was accomplished by randomly shuffling authors across papers 1000 times, recalculating $\alpha$, and taking the median. We then calculated $\alpha'$ by subtracting the null expectation for $\alpha$ from the observed value. We also used the null-simulated $\alpha$ values to calculate a two-tailed p-value for the observed value of $\alpha$; the p-value was defined as the proportion of null simulations for which $|\alpha_{null}| > |\alpha_{obs}|$. We applied false discovery rate correction to each set of p-values to account for multiple testing [@Benjamini_1995].

@@

## Minimising the Wahlund effect: discipline and time period
As discussed in the Introduction, the Wahlund effect can give the appearance of gender homophily (reflected in our study as $\alpha'$ > 0) whenever one lumps together different ???subpopulations??? of academic papers that differ in their author gender ratios (Figure 1).

@@

## Statistical analysis
Previous authors (e.g. @Bonham_2017) have hypothesised that senior scientists preferentially recruit staff and students of the same gender, and/or that junior researchers preferentially select same-gendered mentors. In the majority of disciplines, authorship conventions mean that the first-listed author is often an early-career researcher, while the author listed last is more likely to be a senior researcher leading a research team [@Wren_2007]. Assuming that senior researchers are the main drivers of homophily and that there are enough papers with three or more authors, we predict that the last author???s gender will be the strongest predictor of the remaining authors??? genders (i.e. the gender of the last author will be more salient than that of the first author, or any other authorship position). This is because the first author???s gender would simply be an imperfect correlate of the true causal effect, while the last author???s gender would be the causal effect itself.

@@

$\alpha'$ was significantly higher in the sample from 2015-16 relative to 2005-6, but the difference in means was small (Figure S3; Effect of the fixed factor ???time period??? in a linear mixed model of the data for all author positions: Cohen???s $d$ = 0.09??0.04, $t_{948}$ = 2.51, p = 0.012).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (Figure S4; Effect of the fixed factor ???Authorship position??? in a linear mixed model: Cohen???s $d$ = 0.07??0.02, $t_{1988}$ = 4.48, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors??? genders than the gender of the last author, i.e. the opposite of our prediction based on the hypothesis that senior scientists are causally responsible for homophily.

## Possible differences in the strength of homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was 0.12??0.02 for Urology journals, and 0.03??0.01 for Veterinary Medicine journals; \autoref{alpha_histograms}, Table S3). However, there was little evidence for strong differences in $\alpha'$ between disciplines: the random factor ???discipline??? explained at most 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). This implies that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

There was no indication that journals publishing on a wide range of topics have higher $\alpha'$ values than more specialised journals, due to the Wahlund effect. For example, the journal category ???Multidisciplinary??? - which includes journals like _PLoS ONE_, _Nature_, _Science_, and _PNAS_ - did not have notably elevated $\alpha'$ (\autoref{alpha_histograms}). This result suggests that our estimates of homophily, and estimates from earlier studies, are not notably inflated by the presence of disparate research topics (with variable author gender ratios) being published within individual journals.

@@

# Discussion
We found evidence that researchers tend to publish with same-gendered coauthors, despite implementing stringent controls for ???spurious??? homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our data therefore reaffirm earlier conclusions [@Ferber_1980; @McDowell_1992; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Zettler_2017; @Teele_2017] that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Many journals had $\alpha'$ values over 0.1, meaning that the gender ratio of the average man's coauthors is skewed towards men by >10%, relative to the gender ratio of a woman's coauthors. Few journals had $\alpha'$ values below zero, and fewer still had significantly negative $\alpha'$ values, suggesting that it is comparatively rare for researchers to preferentially select opposite-gendered collaborators.

We hypothesized that disciplines with strong gender-biases (e.g. physics and computer science for men and nursing, midwifery for women [@Holman_2018]) would display a higher level of gender-homophily due to the increased difficulty to find an opposite-gender co-author. However, we found no evidence that gender homophily is restricted to particular disciplines: instead, there was evidence of homophily across many diverse research disciplines (FIG). Gender homophily is thus a global concern rather than an artifact of the existing gender gap (REF). Additionally, we found that gender homophily was slightly stronger in research disciplines with a female-biased gender ratio, as compared to male-biased or gender-balanced fields (whether it is in the first or last author position). One interpretation of this result is that men are more likely to seek out other men in fields in which men are in the minority, relative to the homophily shown by women in fields in which women are in the minority. This male-driven homophily could be, in part, due to the shortage of opportunities for women to participate in international research projects (@Elsevier report),  or international conferences (REF) and therefore less likely to collaborate across gender or fields.

We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. This result suggests that recent increases in the number of women working in STEMM, as well as recent efforts to recruit, retain, and celebrate women in STEMM, have co-occurred with rise in the frequency of same-gender collaborations, not a fall. We speculate that this trend might result, at least in part, from the increasing number of women working in senior positions in STEMM over the last 10 years [@Bendels_2018; @Long_2015; @McKenzie_2017]. Coupled with an ever increase proportion of women in the first author position [@West_2013], this trend leads to an increase in same-sex collaboration. Because many co-authorships involve a junior and a senior researcher (e.g. a student and professor), and the gender ratio is typically more skewed among senior staff [@Shaw_2012, @West_2013], there will be more opposite-gender collaborations than expected if the gender ratio were the same at both career stages. Thus, the observed increase in homophily might partially be due to the greater numbers of women working in leading roles. However, gender homophily during collaboration and co-authoring of scientific manuscript may also impact negatively women careers as there is still a smaller proportion of women in science and thus, making it harder to find female collaborators (Bentley and Adamson 2003).

We also found that journals with a high impact factor (relative to other journals in the same research discipline) tended to have weaker gender homophily than did low-impact journals. If one accepts the contentious claim that papers published in high-impact journals tend to be of higher quality than those in low-impact journals [e.g. @Garfield_2006], then this result provides correlational support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Campbell_2013]. However, journal impact factors tend to be negatively associated with the proportion of women [@Holman_2018. REF]. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor (standardised by research discipline), as predicted if mixed-gender teams produce better research.

Academic publishing is one of the most (if not the most) important tool for a researcher to display their work, and is frequently used as a success indicator for being promoted, hired as a faculty member, or being invited as plenary speaker. Our finding that researchers are more likely to publish with same-gendered coauthors implies that co-authorship listing is not a random process, but instead creates additional gender imbalance. Homophily could make it harder for women to find collaborators in disciplines where women are in the minority [i.e. almost all STEMM fields; @Holman_2018]. Because women make up for as little as 30% of the STEMM scientist community [@Holman_2018], they are therefore less likely to be picked as coauthors, and, as a result, gender homophily creates further gender imbalance and disadvantages for women. Men tend to publish more in high impact journals as first author [@West_2013], and generate more citations than papers lead by first author women [@Long_1992; @Lariviere_2013]. And as a consequence, gender homophily is also likely to contribute to lower publication rates for women [@McDowell_1992], and lower scientific success (REF).

Although the proportion of women in science has significantly increased over the last decades (REF), women still tend to leave their academic careers behind considerably more than their male counterpart (D??barre et al 2018). Lack of scientific recognition (Lincolm et al- 2012), research funding (Bozeman and Corley 2004), travel opportunities, invitation to speak at scientific events (Isbell et al. 2012; Schroeder et al- 2013; Klein et al- 2017), discrimination during hiring (Moss-Racusin et al- 2012; Reuben et al. 2014), when writing a recommendation letter (Trix and Psenka 2003; Schmader et al- 2007), or personal reasons (e.g. child rearing or family ties (Zimmer et al- 2006)), all in all this concentration of small disadvantages drives women out of science. Whether their decision to leave academia is deliberate or not, it will influence the future of the scientific community, as mentoring, networking, and supporting younger women scientists, are especially important for women in academia.

Why does gender homophily exist in science? Collaboration often leads to co-authorship publications, citations and additional professional recognitions, therefore first author should always be considering collaborators that are most likely to enhance their productivity and in the long term also their academic success, rather than pick co-authors according to their gender. First of all, it is likely that decades ago, the probability to meet a woman colleague working on the same area of expertise was slimmer than now. However, the number of women in STEM is increasing, and it is not likely to be an obstacle to collaboration nowadays.

# Acknowledgements
We are grateful to XXXX.This work was supported by the Academy of Finland (284666 to the Centre of Excellence in Biological Interactions).
